Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-09 paper/main.tex
fixes in proofs, related to cycles

	Causal discovery has received significant attention in the Directed
	Acyclic Graphs (DAGs) literature, leading to the development of
	numerous automated algorithms for learning DAGs from data. However,
	their adoption in applied domains remain limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to make obvious errors and output
	Markov Equivalence Classes (MECs) rather than a DAG. To assist
	researchers in constructing DAGs manually, we propose an iterative DAG
	construction approach that combines domain knowledge with data-driven
	insights. Our method leverages Conditional Independence (CI) testing to
	iteratively identify variable pairs where an edge is either missing or
	superfluous. Based on this information, we can choose to add missing
	edges with appropriate orientation based on domain knowledge or remove
	unnecessary ones. We also give a method to rank potential new edges
	based on their impact on the overall model fit. Empirical results show
	that this iterative approach achieves performance comparable to
	automated algorithms if the orientation of new edges is correctly
	determined in at least one out of two cases. Additionally, we show
	that, in the absence of an expert, a Large Language Model (LLM) can be
	used to determine edge orientations. We provide an implementation of
	our approach as both a webtool: <redacted for review> and a Python
	package <redacted for review>.

@@

Our approach to causal structure learning combines domain knowledge with data-driven
insights in a manner that is based on the following base assumptions: 
(1) Domain experts are good at determining causal directions between variables if 
there is a clear causal direction between them. (2) Domain experts are less good
at identifying cases where there is no causal relationship between the variables,
since \emph{potential} causal relationships can often be argued for anyway.
(3) Domain experts are even less good at distinguishing direct from indirect 
effects, since the question whether a direct effect exists between two variables
depends on all other variables present in the graph.

@@

That is, our ancestral oracles never distinguish whether the relationships
between variables is direct or indirect; and the weak ancestral oracle is even
unable to detect cases where there is no cause-effect relationship between two
variables. Using such oracles for DAG construction therefore inherently risks
that superfluous edges are created. In our approach, such superfluous edges
will be detected and removed by utilizing the data.

To model how our learning procedure interacts with the data, we assume that we
have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X \ci_G Y \mid \mathbf{Z}$. These are
the standard oracles considered in constraint-based structure learning
algorithms, such as the PC algorithm \cite{Spirtes2001}. Based on this oracle,
we now define two procedures that iteratively change a current DAG structure.

Our structure learning algorithm is based on two core procedures.
\textsc{Expand} (Algorithm~\ref{algo:expand}) uses conditional independence
information to search for missing edges in the graph and then uses domain
knowledge to orient them. \textsc{Prune} (Algorithm~\ref{algo:prune}) uses
conditional independence information to remove superfluous edges from the
graph.

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B=\emptyset$,$k=\infty$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
            $L \gets L \cup \mathcal{A}_G(X, Y)$ \;
        }
	\If{$|L|\geq k$}{\KwGoTo 12\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding edges based on data and domain knowledge}
\label{algo:expand}
\end{algorithm}

@@

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a strong ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G)=G^+$.
\end{proposition}

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a weak ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G)\supseteq G^+$.
\end{proposition}

@@

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative DAG construction method that
combines domain knowledge with data-driven insights. Our approach iteratively
identifies pair of variables with a missing edge in the DAG, and asks for the
orientation of the edge from an expert. We also propose a method to rank these
potential new edges based on their impact on the model fit. Empirical results
show that this iterative model construction approach achieves performance
comparable to automated algorithms if we are able to correctly decide the
orientation of new edges in one out of three cases. Additionally, we
demonstrate that, in the absence of an expert, a Large Language Model (LLM) can
be used to determine edge orientations. We provide the implementation of our
approach in both a web tool at: <redacted for review> and a Python package
<redacted for review>.

@@

Our approach to causal structure learning combines domain knowledge with data-driven
insights in a manner that is based on the following considerations: 
(1) A domain experts is possibly good at determining causal directions between variables if 
there is a clear causal direction between them. (2) A domain expert may be less good
at identifying cases where there is no causal relationship between the variables,
since \emph{potential} causal relationships can often be argued for anyway.
(3) Many domain experts could struggle to distinguish direct from indirect 
effects, since the presence of a direct effect between two variables
of interest depends on all other variables present in the graph.

@@

That is, our ancestral oracles only provide information on the direction of cause-effect
relationships, and do not distinguish between direct and indirect effects. 
The weak ancestral oracle is even unable to detect cases where there is no cause-effect
relationship between two variables. Using such oracles for DAG construction therefore 
inherently risks that superfluous edges are created. In our approach, such superfluous edges
will be detected and removed by utilizing the data.

To model how our learning procedure interacts with the data, we assume that we
additionally have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X$ and $Y$ are $d$-separated by $\mathbf{Z}$ in 
$G$. These are the standard oracles considered in constraint-based structure learning
algorithms, such as the PC algorithm \cite{Spirtes2001}.

Based on these two oracles, we now define two core procedures that we will use to 
iteratively change a current DAG structure. The procedure 
\textsc{Expand} (Algorithm~\ref{algo:expand}) uses conditional independence
information to search for missing edges in the graph and then uses domain
knowledge to orient them. The procedure \textsc{Prune} (Algorithm~\ref{algo:prune}) uses
conditional independence information to remove superfluous edges from the
graph.

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $e \gets \mathcal{A}_G(X, Y)$ \;
	    \uIf{ $(V,E \cup L \cup e)$ is acyclic }{
            	$L \gets L \cup  \{ e \} $\;
	    } \Else {
            	$L \gets L \cup  \{ \textrm{reverse}(e) \} $\;
	    }
        }
	\If{$|L|\geq k$}{\KwGoTo 12\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

@@

Let $G^+$ denote the transitive closure of $G$, that is, the graph in which 
there is a direct edge $X \to Y$ whenever $X \in \textrm{An}_G(Y)$. The transitive
closure is related to the result of $\textsc{Expand}$ by the following two 
easily proved facts.

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a strong ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G, \emptyset, \infty)=G^+$.
\end{proposition}

\begin{proof}
Since we start from an empty graph, every pair of vertices $X, Y$ where
 $X \in \tetxrm{An}_G(Y)$ is d-separated by the empty set and is
 connected by the edge $X \to Y$. No other edges are added. Therefore, 
the result is $G^+$. 
\end{proof}

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a weak ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G, \emptyset, \infty)
\supseteq G^+$.
\end{proposition}

\begin{proof}
Every pair of vertices $X, Y$ that are $d$-connected in $G$ are now connected by an edge 
during $\textsc{Expand}$. For $X \in \tetxrm{An}_G(Y)$, the edge is 
$X \to Y$ just like when using a strong ancestral oracle; since only correctly oriented
edges are added between ancestor-descendant pairs, such additions never create cycles.
For $X, Y$ that are only $d$-connected throug common ancestors, an arbitrary 
edge is added in a manner that preserves acyclicity. The result is a supergraph
of $G^+$.
\end{proof}

@@

Unlike our ancestral oracles, the pruning operation is quite effective at distinguishing
direct and indirect effects, as shown by the following result.
